Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-26 paper/paper.tex
turn off comments. fix proof.
Extend the eta section

\commentstrue         %% toggle comments here
\extendedfalse
\anonymoustrue

@@

Note that the specification of our logical relation does not require
any changes. We can continue using $\ottkw{ne}$ and $\ottkw{nf}$ to
represent $\beta$-neutral and normal forms. By reproving the
fundamental lemma, we learn that every well-typed term as a
$\beta$-normal form. Lemma~\ref{lemma:parnenf} still holds with our
extended parallel reduction. In the presence of the $\eta$ reduction
rule, Lemma~\ref{lemma:parnenf} tells us that $\eta$ reduction
preserves $\beta$-normal forms (i.e. does not produce new
$\beta$-redexes). Furthermore, since the $\eta$ reduction rule for
functions strictly decreases the size of the term, the existence of
$\beta\eta$ normal form trivially follows.
\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}
Due to some technical
details, our proof about the existence of $\beta\eta$ form does not
immediately induce a decision procedure for type conversion in
\lang{}. We discuss this issue and its workaround in
Section~\ref{sec:conversionalgo}.

\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ l |  c  | c | c }
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    % Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  68 & - & - \\
    Renaming & 46 & -  & - \\
    Untyped reduction & 349 & - & - \\
    Neutral and normal forms & N/A & 269 & N/A \\
    Logical relation & 331 & 426 & N/A \\
    Semantic typing and soundness & 169 & 199 & N/A \\
    Syntactic soundness & N/A & N/A  & 660 \\
    \hline
    Total & 963 & 1357 & 1123 \\
  \end{tabular}
  \end{minipage}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

@@

\citet{anand2014towards} mechanizes the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
extensional type theory with features such as dependent functions,
inductive types, and a full universe hierarchy. \citet{nbeincoq}
mechanizes the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
\citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. However, instead of explaining the inductive
logical relation as a convoluted workaround to the lack of
induction-recursion in Coq, we give a self-contained explanation of
our logical relation.

@@

With untyped conversion,
we sidestep the relational, Kripke-style logical relation found in
other mechanized proofs. \scw{Need to define Kripke-style. Also the other
proofs need Kripke style because they are defining typed relations, not untyped
relations. }
However, our early dependence on confluence
before the fundamental theorem is established can be alarming.
In a system with type-directed reduction,
confluence is not immediately available because it
depends on $\Pi$-injectivity, which is usually only proven after the
fundamental theorem.\scw{confluence depends on Pi injectivity? I thought
it was only needed for subject reduction}\yl{it's
transitive. Confluence depends on subject reduction, which in turn
depends on pi injectivity. Maybe it's worth spelling out the details}
Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
problem that allow us to recover confluence property independently
from the logical relation. \citet{siles2012pure} generalize the
notion of Type Parallel One Step Reduction from \citet{adams2006pure}
to syntactically prove $\Pi$-injectivity for arbitrary Pure Type
Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
equational theory, thus allowing subject reduction to be proven
independently from confluence. By adopting these techniques that allow
us to derive confluence early even for systems with type-directed
reduction, we believe our proof technique can significantly shorten
the existing logical relation proofs for systems with typed judgmental
equality and more complex $\eta$ laws. We leave that as part of our
future work.
% Therefore, we believe that even in a
% system where type-directed reduction is required (e.g. a system with
% the unit $\eta$-law) in the logical relation, the proof can still be carried
% out in a structure similar to the one we have presented.
\scw{Not sure that I understand this paragraph}
\yl{Reworded slightly to emphasize it's future work that we haven't
  done and it is speculative}

>>>>

\commentsfalse         %% toggle comments here
\extendedfalse
\anonymoustrue

@@

Note that the specification of our logical relation does not require
any changes. We can continue using $\ottkw{ne}$ and $\ottkw{nf}$ to
represent $\beta$-neutral and normal forms. By reproving the
fundamental lemma, we learn that every well-typed term as a
$\beta$-normal form. Lemma~\ref{lemma:parnenf} still holds with our
extended parallel reduction. In the presence of the $\eta$ reduction
rule, Lemma~\ref{lemma:parnenf} tells us that $\eta$ reduction
preserves $\beta$-normal forms (i.e. does not produce new
$\beta$-redexes). Furthermore, since the $\eta$ reduction rule for
functions strictly decreases the size of the term, the existence of
$\beta\eta$ normal form trivially follows.
\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}

In \citet{nbeincoq,decagda,martin-lof-a-la-coq}, a relational model is
needed to justify the $\eta$-law for functions. Instead, we prove
confluence for our parallel reduction extended with $\eta$-law for functions
and continue using a logical predicate (i.e. a 1-ary logical relation) to justify
$\eta$-law as part of our equational theory.
A well-known issue with our approach is the failure of syntactic
confluence when the lambda term contains type annotations. A simple
counterexample is $[[\- B ((\- A a) {up 1} var 0)]]$; depending on
whether \rref{P-AbsEta} is performed on the whole term or
\rref{P-AppAbs} is used on the inner $\beta$ redex, we end up with the
terms $[[\- B a]]$ or $[[\- A a]]$, where $[[A]]$ and $[[B]]$ are not
necessarily syntactically equal terms. \citet{choudhury:ddc} resolve
this problem by stating their confluence result in terms of an
equivalence relation that quotients out parts of the terms that are
computationally irrelevant; the annotations of lambda terms are
ignored since the behavior of a lambda term is not affected by its
type annotation. We believe the same approach is applicable to our
proof if we add type annotations to our lambda terms.

The bigger issue is extensions such as $\eta$-laws for unit and
products. Surjective pairing, for example, is an equational law that
is known to cause trouble since it is not confluent for untyped lambda
terms~\citet{KLOP198997}. The relational, type-annotated, and Kripke-style models from
\citet{nbeincoq,decagda,martin-lof-a-la-coq} can be more easily
extended to support these rules.
We note, however, that the issue with $\eta$ rules is not exclusive to dependently
typed languages and has been studied in more limited languages that
are either simply
typed~\citep{pierce2004advanced,pfenning1997computation} or
dependently typed but without large
elimination~\citep{harper2005equivalence,
abel2005untypedconvsurjective}. Common workarounds include
type-directed conversion and shifting the focus to obtaining
$\eta$-long forms~\citet{Abel12}.
We believe our simple proof will help readers
see how techniques from these simpler settings generalize to fully
dependently typed languages and better understand the more complex
mechanized developments.

Finally, due to some technical
details, our mechanized proof about the existence of $\beta\eta$ form does not
immediately induce a decision procedure for type conversion in
\lang{}. We discuss this issue and its workaround in
Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ l |  c  | c | c }
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    % Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  68 & - & - \\
    Renaming & 46 & -  & - \\
    Untyped reduction & 349 & - & - \\
    Neutral and normal forms & N/A & 269 & N/A \\
    Logical relation & 331 & 426 & N/A \\
    Semantic typing and soundness & 169 & 199 & N/A \\
    Syntactic soundness & N/A & N/A  & 660 \\
    \hline
    Total & 963 & 1357 & 1123 \\
  \end{tabular}
  \end{minipage}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

@@

\citet{anand2014towards} mechanizes the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
extensional type theory with features such as dependent functions,
inductive types, and a full universe hierarchy. \citet{nbeincoq}
mechanizes the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the
only base type from the object language, large elimination is not
supported despite the one predicative universe.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further shows it is possible
to encode a finite universe hierarchy without the use of
either impredicativity or induction-recursion.
% \yl{commented out because I think this paragraph is too subjective}
% However, instead of explaining the inductive
% logical relation as a convoluted workaround to the lack of
% induction-recursion in Coq, we give a self-contained explanation of
% our logical relation.

@@

With untyped conversion,
we sidestep the relational, Kripke-style logical relation found in
other mechanized proofs. \scw{Need to define Kripke-style. Also the other
proofs need Kripke style because they are defining typed relations, not untyped
relations. }
However, our early dependence on confluence
before the fundamental theorem is established can be alarming.
In a system with type-directed reduction,
confluence is not immediately available because it
depends on $\Pi$-injectivity, which is usually only proven after the
fundamental theorem.\scw{confluence depends on Pi injectivity? I thought
it was only needed for subject reduction}\yl{it's
transitive. Confluence depends on subject reduction, which in turn
depends on pi injectivity. Maybe it's worth spelling out the details}
Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
problem that allow us to recover confluence property independently
from the logical relation. \citet{siles2012pure} generalize the
notion of Type Parallel One Step Reduction from \citet{adams2006pure}
to syntactically prove $\Pi$-injectivity for arbitrary Pure Type
Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
equational theory, thus allowing subject reduction to be proven
independently from confluence. By adopting these techniques that allow
us to derive confluence early even for systems with type-directed
reduction, we believe our proof technique can significantly shorten
the existing logical relation proofs for systems with typed
judgmental. We leave that as part of our future work.
% Therefore, we believe that even in a
% system where type-directed reduction is required (e.g. a system with
% the unit $\eta$-law) in the logical relation, the proof can still be carried
% out in a structure similar to the one we have presented.
\scw{Not sure that I understand this paragraph}
\yl{Reworded slightly to emphasize it's future work that we haven't
  done and it is speculative}
